---
title: "Reproducible Research: Peer Assessment 1"
author: "yasser zhang"
date: "Tuesday, January 13, 2015"
output: html_document
---
### Description of the data

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.  

The original dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.  

The data have three variables, namely, steps, date, and interval.  

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
- date: The date on which the measurement was taken in YYYY-MM-DD format
- interval: Identifier for the 5-minute interval in which measurement was taken

### Loading and preprocessing the data

I load the data in the format of data frame and assign it to the variable "activity". Then convert the dataset into tbl class, which is easier to manipulate the data.

```{r}
library(dplyr)
activity <- read.csv("~/activity.csv",header=T)
activity <- tbl_df(activity)
```

### What is mean total number of steps taken per day?

```{r}
total_per_day <- activity %>%
    group_by(date) %>%
    summarise(sum=sum(steps)) %>%
    filter(!is.na(sum))
```
```{r}
hist(total_per_day$sum,xlab="total number of steps taken per day",
     main="Frequency Distribution of total steps per day")
```


### What is the average daily activity pattern?



### Imputing missing values



### Are there differences in activity patterns between weekdays and weekends?