---
title: "Reproducible Research: Peer Assessment 1"
author: "Ning Zhang"
date: "Tuesday, January 13, 2015"
output: html_document
---
### The purpose of the assignment

This assignment is implemented to develop the ability to write a reproducible research report based on a dataset using knitr.

### Description of the data

In this assignment, I use the data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.  

The original dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset. Here is the [download link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).  

The data have three variables, namely, steps, date, and interval.  

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
- date: The date on which the measurement was taken in YYYY-MM-DD format
- interval: Identifier for the 5-minute interval in which measurement was taken

### Loading and preprocessing the data

I load the data in the format of data frame and assign it to the variable "activity". Then convert the dataset into tbl class, which is easier to manipulate the data.

```{r message=FALSE}
library(dplyr)
activity <- read.csv("./activity.csv",header=T)
activity <- tbl_df(activity)
```

### What is mean total number of steps taken per day?

In this step, I want to know how many steps in total this person moves in each day. So I group the data by date and calculate the daily sum of the steps of 5 minute intervals. There are eight days whose values in the sum of daily steps are NA values. I removed these days from the final data.
```{r}
total_per_day <- activity %>%
    group_by(date) %>%
    summarise(sum=sum(steps)) %>%
    filter(!is.na(sum))
```

Below is what the daily total steps the subject takes during the two months look like in histogram chart.

```{r}
library(ggplot2)
hist_total <- ggplot(total_per_day,aes(sum)) + xlim(0,25000)
hist_total + geom_histogram(binwidth=1000,colour = "white",fill = "black",alpha=0.8) +
  xlab("total number of steps taken per day") +
  ylab("Days") + 
  ggtitle("Frequency Distribution of Total Steps Taken Per Day") 
```

With daily total steps calculated, I want to know the mean and median value of daily total steps during the two months. I calculate them and assign them to "mean_per_day" and "median_per_day".  

```{r}
options(scipen=3, digits=2)
mean_per_day <- mean(total_per_day$sum)
median_per_day <- median(total_per_day$sum)
```
```{r}
mean_per_day
median_per_day
```

The numeric difference between the mean total number of steps taken per day,`r mean_per_day`, and its median, `r median_per_day` is around *one* step. The difference is trivial, so there is little skewness in the distribution of total number of steps taken per day across the two months.

### What is the average daily activity pattern?

In this step, I explore how many steps the subject takes at the same time interval in each day during the two measured months. In the dataset, one day of 24 hours or 1,440 minutes is divided into 288 ordered five-minute intervals. The value space of 288 five-minute time intervals are stored at the variable of "interval", which starts with 0, increases by a step of 5, and ends at 2355.

I group the dataset by the variable of "interval", and calculate the average value in each of the 288 intervals. A line plot shown below is drawn based on the mutated data. The plot's x-axis represents five-minute intervals, and y-axis represents the average number of steps taken in a specific time interval across all days.

```{r}
average_per_day <- activity %>%
    group_by(interval) %>%
    summarise(avg = mean(steps,na.rm=T))
avg_step <- ggplot(average_per_day, aes(interval,avg))
avg_step + 
    geom_line(colour="black",size = 2,alpha=0.7) +
    ggtitle("Average Daily Activity Pattern") + 
    xlim(0,2355) +
    xlab("five-minute interval") + 
    ylab("average steps")
```

Which 5-minute interval contains the maximum number of steps?

```{r}
average_per_day[which(average_per_day$avg==max(average_per_day$avg)),]$interval
```

The maximum number of steps happen at the 835 interval, or say, during 8:35 to 8:40 a.m. every day.

### Imputing missing values

The calculation done above does not take NA values into account, simply ruling them out. Let's see if the results may differ after I fix all NA inputs.  

I calculate the total number of missing values in the dataset.

```{r}
row_na <- nrow(activity[is.na(activity),])
row_na
```

There are `r row_na` rows in the data containing missing values.  

How do I deal with all of the NA values?

I replace each NA value with the average number of steps taken in the 5 minute interval where the NA value is taken. And what I get is a new dataset I call "new_activity".

```{r}
new_activity <- activity %>%
    group_by(interval) %>%
    mutate(steps = ifelse(is.na(steps),mean(steps,na.rm=T),steps))
```

Then I re-draw a histogram of the total number of steps taken each day, and .

```{r}
new_total_per_day <- new_activity %>%
  group_by(date) %>%
  summarise(sum=sum(steps))

new_hist_total <- ggplot(new_total_per_day,aes(sum)) + xlim(0,25000)
new_hist_total + geom_histogram(binwidth=1000,colour = "white",fill = "black",alpha=0.8) +
  xlab("total number of steps taken per day") +
  ylab("Frequency") + 
  ggtitle("Frequency Distribution of Total Steps Taken Per Day") 
```

I also re-calculate the mean and median total number of steps taken per day.

```{r}
new_mean_per_day <- mean(new_total_per_day$sum)
new_median_per_day <- median(new_total_per_day$sum)
new_mean_per_day
new_median_per_day
```

The new mean total number of steps taken per day is `r new_mean_per_day`, and the new median value is `r new_median_per_day`. The two values happen to be the same, and have no significant difference from their counterparts which are calculated regardless of NA values.

### Are there differences in activity patterns between weekdays and weekends?

In this step, I want to know if there exist different activity patterns between weekdays and weekends. I create a new factor variable tagging each row of observations with "weekday" or "weekend" in the “new_activity” dataset.

```{r}
weekday  <- c("Monday","Tuesday","Wednesday","Thursday","Friday")
new_activity <- new_activity %>% 
  mutate(day=ifelse(weekdays(as.Date(date)) %in% weekday, "weekday", "weekend")) %>%
  mutate(day = as.factor(day))
```

Then I make two line plots for weekday and weekend of the 5-minute intervals (x-axis) and the average number of steps taken (y-axis).

```{r}
new_average_per_day <- new_activity %>% 
    group_by(interval,day) %>%
    summarise(avg = mean(steps))
week_plot <- ggplot(new_average_per_day, aes(interval,avg))
week_plot + geom_line(aes(color= day),size = 1,alpha=0.65) + 
    geom_point(aes(color= day), alpha = 0.9, size = 1.5) +
    facet_grid(day ~ .) + 
    ggtitle("Average Activity Patterns of Weekdays and Weekends") + 
    xlab("five-minute interval") + 
    ylab("average steps taken")
```

The end.